Reasonably Programmable Literal Notation

### 2018-02-26 icfp18/syntax-icfp18.tex
initial pass over intro done

The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e. the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}. The literal body is constrained by the context-free syntax of the language only in that nested delimiters must be matched (like comments in OCaml). TSLs are closely related to the mechanism proposed in this paper, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
\begin{enumerate}[leftmargin=12px]

@@

  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e. subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping.

@@

  \item[\LEFTcircle] \textbf{Typing}: Each parser is associated with a type, and the generated expansion must necessarily be of that type, so abstract reasoning about the type of the expansion is possible. However, there is no easy way to determine the type expected for each spliced expression. In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason, which is a syntactic layer over all of OCaml's semantics \cite{ocaml-manual}:

@@

\paragraph{Outline} 
% \begin{itemize}
% \item 
We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, and describes the special hygiene conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} adds type functions and an ML-style module system and introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which solve several problems. 
Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work using the rubric of five reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the specific contributions of this paper and a discussion.

@@

Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are given in the supplemental material. Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

%

@@

>>>>

The parser associated with the type \li{KQuery.t} is statically invoked to lex, parse and expand the literal body, i.e. the sequence of characters between the outer delimiters, here \li{`(} and \li{)`}. The literal body is constrained by the context-free syntax of the language only in that nested delimiters must be matched (like comments in OCaml). TSLs are closely related to the mechanism that we will propose, so let us analyze TSLs from the perspective of the five reasoning criteria just outlined.
\begin{enumerate}[leftmargin=12px,nolistsep,noitemsep]

@@

  \item[\LEFTcircle] \textbf{Segmentation}: The parser can splice expressions out of the literal body, but there is no clear way to associate each spliced term with some particular segment (i.e. subsequence) of the literal body, nor is there a guarantee that spliced terms are non-overlapping. As such, there is no way to indicate to the programmer where base language expressions are located within a literal form.

@@

  \item[\LEFTcircle] \textbf{Typing}: The expansion must necessarily be of the associated type, so abstract reasoning about the type of the expansion is straightforward. However, there is no easy way to determine the type expected for each spliced expression. In addition, the prior work considered only a monomorphic, nominally-typed language with local type inference, leaving open a number of problems that came up as we considered integrating TSLs into Reason/OCaml:

@@

\paragraph{Outline} 
% \begin{itemize}
% \item 
We begin in Sec.~\ref{sec:setsms} by introducing TLMs with a detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs}, i.e. TLMs that generate patterns, and describes the special hygiene conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms}  introduces \emph{parametric TLMs} (pTLMs), i.e. TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types. Sec.~\ref{sec:ptsms} also reveals that the \li{expansions require} clause is syntactic sugar for parameterization followed immediately by partial parameter application. 
Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with additional examples that demonstrate both the expressive power of this approach and certain alternative parser implementation strategies that notation providers might consider. 
Sec.~\ref{sec:setsms-formally} presents a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above, in the formally simplest setting. 
Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives the most general variant of the reasoning principles theorem. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of five reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of future work.

@@

Due to the space limitations, certain technical details and extensions to the calculi presented in this paper, as well as detailed proofs, are available in the supplemental material. % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

\newcommand{\moreExamplesSec}{Additional Examples}
\section{\protect\moreExamplesSec}
\label{sec:more-examples}

\subsection{HTML}

\subsection{Quotation}

\subsection{More Examples, Briefly}
%

@@

\newcommand{\pTLMsFormallySec}{Parametric TLMs, Formally}
\section{\protect\pTLMsFormallySec}
\label{sec:ptlms-formally}

% \begin{figure}[p]
% \[\begin{array}{llcl}
% \mathsf{UMType} & \urho & ::= 
% %& \autype{\utau} 
% & \utau & \text{type annotation}\\
% % &&
% %& \aualltypes{\ut}{\urho} 
% % & \alltypes{\ut}{\urho} & \text{type parameterization}\\
% &&
% %& \auallmods{\usigma}{\uX}{\urho} 
% & \allmods{\uX}{\usigma}{\urho} & \text{module parameterization}\\
% \mathsf{UMExp} & \uepsilon & ::= 
% %& \abindref{\tsmv} 
% & \tsmv & \text{TLM identifier reference}\\
% % &&
% %& \auabstype{\ut}{\uepsilon} 
% % & \abstype{\ut}{\uepsilon} & \text{type abstraction}\\
% &&
% %& \auabsmod{\usigma}{\uX}{\uepsilon} 
% & \absmod{\uX}{\usigma}{\uepsilon} & \text{module abstraction}\\
% % &&
% %& \auaptype{\utau}{\uepsilon} 
% % & \aptype{\uepsilon}{\utau} & \text{type application}\\
% &&
% %& \auapmod{\uM}{\uepsilon} 
% & \apmod{\uepsilon}{\uX} & \text{module application}\ECC
% \end{array}
% \]
% \caption{Syntax of unexpanded TLM types and expressions.}
% \label{fig:P-macro-expressions-types-u}
% \end{figure}

The main conceptual difference between $\miniVersePat$ and $\miniVerseParam$ is that $\miniVerseParam$ introduces the notion of unexpanded and expanded TLM expressions and types, as shown in Fig. \ref{fig:P-macro-expressions-types}.

\begin{figure}[h]
\small
\begin{minipage}{0.38\textwidth}
$\arraycolsep=3pt\begin{array}{llcl}
\mathsf{UMType} & \urho & ::= & \utau ~\vert~ \allmods{\uX}{\usigma}{\urho}\\
\mathsf{UMExp} & \uepsilon & ::= & \tsmv ~\vert~ \absmod{\uX}{\usigma}{\uepsilon}~\vert~ \apmod{\uepsilon}{\uX}
\end{array}$
\end{minipage}
\begin{minipage}{0.6\textwidth}
$\arraycolsep=3pt\begin{array}{llcl}
\mathsf{MType} & \rho & ::= & \aetype{\tau} ~\vert~ \aeallmods{\sigma}{X}{\rho} \\
\mathsf{MExp} & \epsilon & ::= & \adefref{a} ~\vert~ \aeabsmod{\sigma}{X}{\epsilon} ~\vert~\aeapmod{X}{\epsilon} 
\end{array}$
\end{minipage}
\caption{Syntax of unexpanded and expanded TLM types and expressions in $\miniVerseParam$}
\label{fig:P-macro-expressions-types}
\end{figure}
The TLM type $\aeallmods{\sigma}{X}{\rho}$ classifies TLM expressions that have one module parameter matching $\sigma$. For simplicity, we formalize only module parameters. Type parameters can be expressed as module parameters having exactly one abstract type member.

The calculus enjoys metatheoretic properties analagous to those described in Sec. \ref{sec:s-metatheory}, modified to account for the presence of modules, kinds and parameterization. The following theorem establishes the abstract reasoning principles available when applying a parametric expression TLM. The clauses are directly analagous to those of Theorem \ref{thm:setlm-reasoning}, so for reasons of space we do not repeat the inline descriptions. The \textbf{Kinding} clauses can be understood by analogy to the \textbf{Typing} clauses. The details of parametric pattern TLMs (ppTLMs) are analagous (see supplement.)
\vspace{-3px}
\begin{theorem}[peTLM Reasoning Principles]
If $\expandsP{\uOmega}{\uPsi}{\uPhi}{\utsmap{\uepsilon}{b}}{e}{\tau}$ then:
\begin{enumerate}[nolistsep,noitemsep]
  \item $\uOmega=\uOmegaEx{\uD}{\uG}{\uMctx}{\Omega_\text{app}}$
  \item $\uPsi=\uAS{\uA}{\Psi}$
  \item (\textbf{Typing 1}) $\tsmexpExpandsExp{\uOmega}{\uPsi}{\uepsilon}{\epsilon}{\aetype{\tau}}$ and $\hastypeP{\Omega_\text{app}}{e}{\tau}$
  \item $\tsmexpEvalsExp{\Omega_\text{app}}{\Psi}{\epsilon}{\epsilon_\text{normal}}$
  \item $\tsmdefof{\epsilon_\text{normal}}=a$
  \item $\Psi = \Psi', \petsmdefn{a}{\rho}{\eparse}$
  \item $\encodeBody{b}{\ebody}$
    \item $\evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessE}}{e_\text{pproto}}}$
  \item $\decodePCEExp{e_\text{pproto}}{\pce}$
  \item $\prepce{\Omega_\text{app}}{\Psi}{\pce}{\ce}{\epsilon_\text{normal}}{\aetype{\tau_\text{proto}}}{\omega}{\Omega_\text{params}}$
  \item (\textbf{Segmentation}) $\segOK{\segof{\ce}}{b}$
  \item $\cvalidEP{\Omega_\text{params}}{\esceneP{\omega : \OParams}{\uOmega}{\uPsi}{\uPhi}{b}}{\ce}{e'}{\tau_\text{proto}}$
  \item $e = [\omega]e'$
  \item $\tau = [\omega]\tau_\text{proto}$
  \item $
    \segof{\ce} = \sseq{\acesplicedk{m_i}{n_i}}{\nkind} \cup \sseq{\acesplicedc{m'_i}{n'_i}{\cekappa'_i}}{\ncon} \cup $ \\
     ~~~~$          \sseq{\acesplicede{m''_i}{n''_i}{\ctau_i}}{\nexp}
    $
  \item (\textbf{Kinding 1}) $\sseq{\kExpands{\uOmega}{\parseUKindF{\bsubseq{b}{m_i}{n_i}}}{\kappa_i}}{\nkind}$ and \\ ~~~~$\sseq{\iskind{\Omega_\text{app}}{\kappa_i}}{\nkind}$
  \item (\textbf{Kinding 2}) $\sseq{\cvalidK{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\cekappa'_i}{\kappa'_i}}{\ncon}$ and $\sseq{\iskind{\Omega_\text{app}}{[\omega]\kappa'_i}}{\ncon}$
  \item (\textbf{Kinding 3}) $\sseq{\cExpands{\uOmega}{\parseUConF{\bsubseq{b}{m'_i}{n'_i}}}{c_i}{[\omega]\kappa'_i}}{\ncon}$ and $\sseq{\haskind{\Omega_\text{app}}{c_i}{[\omega]\kappa'_i}}{\ncon}$
  \item (\textbf{Kinding 4}) $\sseq{\cvalidC{\OParams}{\csceneP{\omega : \OParams}{\uOmega}{b}}{\ctau_i}{\tau_i}{\akty}}{\nexp}$ and $\sseq{\haskind{\Omega_\text{app}}{[\omega]\tau_i}{\akty}}{\nexp}$
  \item (\textbf{Typing 2}) $\sseq{\expandsP{\uOmega}{\uPsi}{\uPhi}{\parseUExpF{\bsubseq{b}{m''_i}{n''_i}}}{e_i}{[\omega]\tau_i}}{\nexp}$ and $\sseq{\hastypeP{\Omega_\text{app}}{e_i}{[\omega]\tau_i}}{\nexp}$
  \item (\textbf{Capture Avoidance}) $e = [\sseq{\kappa_i/k_i}{\nkind}, \sseq{c_i/u_i}{\ncon}, \sseq{e_i/x_i}{\nexp}, \omega]e''$ for some $e''$ and fresh $\sseq{k_i}{\nkind}$ and fresh $\sseq{u_i}{\ncon}$ and fresh $\sseq{x_i}{\nexp}$
  \item (\textbf{Context Independence}) $\mathsf{fv}(e'') \subset \sseq{k_i}{\nkind} \cup \sseq{u_i}{\ncon} \cup \sseq{x_i}{\nexp} \cup \domof{\OParams}$
  % $\hastypeP{\sseq{\Khyp{k_i}}{\nkind} \cup \sseq{u_i :: [\omega]\kappa'_i}{\ncon} \cup \sseq{x_i : [\omega]\tau_i}{\nexp}}{[\omega]e''}{\tau}$\todo{maybe restate this in terms of free variables of e'' here and elsewhere, because context isn't technically well-formed here?}
\end{enumerate}
\end{theorem}

\vspace{-6px}
